---
title: "STA 138 Final Project"
author: "Zhuoya Wang"
date: "2023-12-04"
output:
  word_document: default
  pdf_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align='center', message = FALSE, warning = FALSE)
```


```{r}
library(ggplot2)
library(dplyr)
#install.packages("ggpubr")
library(ggpubr)
library(MASS)
library(reshape2)
#install.packages("reshape")
library(reshape)
library(data.table)
```



```{r}
### raw data
byss <- read.csv("~/Desktop/Fall 2023/138/final project/Byssinosis.csv")
n <- nrow(byss)
countColumns <- c(which(names(byss) == "Byssinosis"),
                  which(names(byss) == "Non.Byssinosis"))
longByss <- rbind(
  cbind(Byss=1,
        byss[rep(1:n, byss[,"Byssinosis"]), -countColumns]),
  cbind(Byss=0,
        byss[rep(1:n, byss[,"Non.Byssinosis"]), -countColumns])
)
row.names(longByss) <- 1:nrow(longByss)
longByss$Workspace = as.factor(longByss$Workspace)
```



## Data Visualization
```{r}
### Employment years & Y-- PLot 1

ggplot(longByss, aes(x = longByss$Employment, fill = factor(longByss$Byss))) +
  geom_bar(position = position_dodge(width = 0.9), stat = "count") +
  geom_text(stat = 'count', aes(label = ..count..), 
            vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Byssinosis Frequency by employment years",
       x = "Employment Years",
       y = "Frequency")

##### For positive Byss

byssinosis_count <- longByss %>%
  filter(Byss == 1) %>%
  group_by(Employment) %>%
  summarise(Frequency = n())

# Now create the horizontal bar plot
ggplot(byssinosis_count, aes(x = Frequency, y = Employment)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Frequency of Positive Byssinosis by Employment Years",
       x = "Frequency of Positive Byssinosis",
       y = "Employment Years") 

```



```{r}
### Smoking & Byss --plot 2

ggplot(longByss, aes(x = longByss$Smoking, fill = factor(longByss$Byss))) +
  geom_bar(position = position_dodge(width = 0.9), stat = "count") +
  geom_text(stat = 'count', aes(label = ..count..), 
            vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Byssinosis Frequency by smoking", x = "Smoking or not",
       y = "Frequency")
```

```{r}
#### Sex & Byss --- plot3
ggplot(longByss, aes(x = longByss$Sex, fill = factor(longByss$Byss))) +
  geom_bar(position = position_dodge(width = 0.9), stat = "count") +
  geom_text(stat = 'count', aes(label = ..count..), 
            vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Byssinosis Frequency by Sex", x = "Sex(Female & Male)",
       y = "Frequency")
```

```{r}
#### Race & Byss --- plot4
ggplot(longByss, aes(x = longByss$Race, fill = factor(longByss$Byss))) +
  geom_bar(position = position_dodge(width = 0.9), stat = "count") +
  geom_text(stat = 'count', aes(label = ..count..), 
            vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Byssinosis Frequency by Race", x = "Race (Others & White)",
       y = "Frequency")
```

```{r}
#### Race & Byss --- plot5
ggplot(longByss, aes(x = longByss$Workspace, fill = factor(longByss$Byss))) +
  geom_bar(position = position_dodge(width = 0.9), stat = "count") +
  geom_text(stat = 'count', aes(label = ..count..), 
            vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Byssinosis Frequency by Workspace", 
       x = "Workspace[1 (most dusty), 2 (less dusty), 3 (least dusty)]",
       y = "Frequency")
```



```{r}
empty_model  = glm(longByss$Byss ~ 1, data = longByss, family = binomial())
summary(empty_model)
BIC(empty_model)
```


```{r}
### Full Model
full_model = glm(longByss$Byss ~ ., data = longByss, family = binomial())
summary(full_model)
```

####AIC
```{r}
library(MASS)
n_new = nrow(longByss)
AIC_model = stepAIC(full_model, k =2)
```

##### 1201.46 for AIC with model "Employment + Smoking + Workspace"



```{r}
BIC_model = stepAIC(full_model, k = log(n_new))
```

#### lowest BIC is 1237.78 with "Smoking + Workspace"




####AIC model selection with direction
```{r}
forward.model.AIC = stepAIC(empty_model, 
              scope = list(lower = empty_model, upper= full_model),k = 2,
              direction = "forward", trace = FALSE)

backward.model.AIC = stepAIC(full_model, 
              scope = list(lower = empty_model, upper= full_model), 
              k = 2,direction = "backward",trace = FALSE)

both.model.AIC = stepAIC(full_model, 
            scope = list(lower = empty_model, upper= full_model), 
            k = 2,direction = "both",trace = FALSE)

summary(forward.model.AIC)
summary(backward.model.AIC)
summary(both.model.AIC)

forward.model.AIC$aic
backward.model.AIC$aic
both.model.AIC$aic


forward.model.AIC$coefficients
backward.model.AIC$coefficients
both.model.AIC$coefficients

anova(forward.model.AIC,backward.model.AIC, both.model.AIC )

```

##### all three direction choose "Workspace + Smoking + Employment" w/o interaction 


###BIC selection

```{r}
forward.model.BIC = stepAIC(empty_model, 
                    scope = list(lower = empty_model, upper= full_model),
                    k =log(n_new) ,direction = "forward", trace = FALSE)

backward.model.BIC = stepAIC(full_model, 
              scope = list(lower = empty_model, upper= full_model),
              k =log(n_new),direction = "backward",trace = FALSE)

both.model.BIC = stepAIC(full_model, 
                  scope = list(lower = empty_model, upper= full_model),
                  k =log(n_new),direction = "both",trace = FALSE)

summary(forward.model.BIC)
summary(backward.model.BIC)
summary(both.model.BIC)

forward.model.BIC$aic
backward.model.BIC$aic
both.model.BIC$aic


forward.model.BIC$coefficients
backward.model.BIC$coefficients
both.model.BIC$coefficients

anova(forward.model.BIC,backward.model.BIC, both.model.BIC )
```
### BIC choose to use "Workspace + Smoking"



### with interaction model

```{r}
in_full_model <- glm(longByss$Byss~.^2,data=longByss, 
                 family = binomial())
summary(in_full_model)

```



### model selection with new full model

```{r}
in_forward.model.AIC = stepAIC(empty_model, 
            scope = list(lower = empty_model, upper= in_full_model),k = 2,
            direction = "forward", trace = FALSE)

in_backward.model.AIC = stepAIC(in_full_model, 
                scope = list(lower = empty_model, upper= in_full_model), 
                k = 2,direction = "backward",trace = FALSE)

in_both.model.AIC = stepAIC(in_full_model,
      scope = list(lower = empty_model, upper= in_full_model), 
      k = 2,direction = "both",trace = FALSE)

summary(in_forward.model.AIC)
summary(in_backward.model.AIC)
summary(in_both.model.AIC)

in_forward.model.AIC$aic
in_backward.model.AIC$aic
in_both.model.AIC$aic
```
##  "Workspace + Smoking + Employment +  Workspace:Smoking" has the lower AIC which is 1200.771




###BIC
```{r}
in_forward.model.BIC = stepAIC(empty_model, 
                          scope = list(lower = empty_model, upper= in_full_model),
                         k = log(n_new),direction = "forward", trace = FALSE)

in_backward.model.BIC = stepAIC(in_full_model, scope = list(lower = empty_model, 
    upper= in_full_model), k = log(n_new),direction = "backward",trace = FALSE)

in_both.model.BIC = stepAIC(in_full_model, scope = list(lower = empty_model,
      upper= in_full_model), k = log(n_new),direction = "both",trace = FALSE)

summary(in_forward.model.BIC)
summary(in_backward.model.BIC)
summary(in_both.model.BIC)

in_forward.model.BIC$aic
in_backward.model.BIC$aic
in_both.model.BIC$aic


anova(in_forward.model.AIC, in_backward.model.AIC,in_both.model.AIC )
anova(in_forward.model.BIC,in_backward.model.BIC,in_both.model.BIC )
```


###### deviance plot

```{r}

BIC_choice = glm(formula = longByss$Byss ~ Smoking + Workspace, family = binomial(), 
    data = longByss)

AIC_choice = glm(formula = longByss$Byss ~ Workspace + Smoking + Employment + 
    Workspace:Smoking, family = binomial(), data = longByss)


library(ggplot2)
deviance_values <- c(deviance(AIC_choice), deviance(BIC_choice))
model_names <- c("AIC_choice", "BIC_choice")

plot_data <- data.frame(Model = model_names, Deviance = deviance_values)

ggplot(plot_data, aes(x = Model, y = Deviance)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  theme_minimal() +
  labs(title = "Model Comparison by Deviance", x = "Model", y = "Deviance")

```

```{r, ref.label=knitr::all_labels(), eval = F, echo = T}

```


